---
title: "Assignment 2: Coding Basics"
author: "Matthew Vining"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1.
basics_sequence <- seq(1, 100, 4) ### assigning bascis_sequence a array of number 1 through 100 and by increments of 4

#2. 
mean_b <- mean(basics_sequence) 
mean_b
# calculating and assigning the mean of the basics_sequence array

med_b <- median(basics_sequence)
med_b
# calculating and assigning the median of the basics_sequence array

#3.
mean_b > med_b
#setting the conditional statement that the mean of b is greater than the median of b to see if R returns a true or false statement
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
vector_names <- c("Matthew","Nef","Abby","Sitara") # student names character vector
vector_names
vector_testscore <- c(98,80,95,88) # numeric vector of test scores
vector_testscore
vector_passed <- c(TRUE,TRUE,TRUE,TRUE) #logical vector if they passed test or not
vector_passed

test_dataframe <- data.frame(vector_names,vector_testscore,vector_passed) 
test_dataframe
#create a data frame that combines each of the vectors described above

names(test_dataframe) <- c("Name","Test Score","Pass(TRUE)/Fail(FALSE)"); View(test_dataframe) # labeled columns of data
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Unlike a matrix where columns have to have the same data types, columns in a data frame can have different types of data like character vs numeric.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
pass_fail <- function(vector_testscore) {
 ifelse(vector_testscore >= 50, "Pass", "Fail") #log_exp, if TRUE, if FALSE
}

pass_fail <- function(vector_testscore) {
  if(vector_testscore >= 50) {
    "Pass"
  } 
  else {
    "Fail"
  }
}
TestResultMatthew <- pass_fail(98); TestResultMatthew
TestResultNef <- pass_fail(80); TestResultNef
TestResultAbby <- pass_fail(95); TestResultAbby
TestResultSitara <- pass_fail(88); TestResultSitara
# running if else test to see who passes or not
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ifelse worked the best because it was easier to write. The if and else function required a more complex loop and code to create and I found it easier to run into issues than if I use the ifelse code langauge. However, both types worked and gave the same output.
